---
title: "Comprehensive Guide to Bayesian Computation and MCMC Methods"
author: "Advanced Statistical Methods"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: readable
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(ggplot2)
library(MASS)
library(coda)
library(magicaxis)
library(viridis)
set.seed(42)
```

# Introduction to Bayesian Computation

Bayesian computation forms the backbone of modern statistical inference, providing a principled framework for learning from data while incorporating prior knowledge. Unlike frequentist approaches that treat parameters as fixed but unknown quantities, Bayesian methods treat parameters as random variables with probability distributions.

## Core Concepts

### Bayes' Theorem
The fundamental equation governing Bayesian inference:

$$P(\theta|D) = \frac{P(D|\theta)P(\theta)}{P(D)}$$

Where:
- $P(\theta|D)$ = **Posterior**: Our updated belief about parameters after seeing data
- $P(D|\theta)$ = **Likelihood**: Probability of observing data given parameters  
- $P(\theta)$ = **Prior**: Our initial belief about parameters before seeing data
- $P(D)$ = **Evidence/Marginal Likelihood**: Normalizing constant

### The Challenge
The denominator $P(D) = \int P(D|\theta)P(\theta)d\theta$ is often intractable analytically, especially for complex models with multiple parameters. This is where **Markov Chain Monte Carlo (MCMC)** methods become essential.

# Monte Carlo Methods

## Basic Monte Carlo Integration

Monte Carlo methods use random sampling to solve computational problems. For integration:

$$\int_a^b f(x)dx \approx \frac{b-a}{N}\sum_{i=1}^N f(x_i)$$

where $x_i$ are random samples from $\text{Uniform}(a,b)$.

```{r monte_carlo_basic}
# Basic Monte Carlo integration example
monte_carlo_integrate <- function(f, a, b, n_samples = 10000) {
  x <- runif(n_samples, a, b)
  y <- sapply(x, f)
  estimate <- (b - a) * mean(y)
  se <- (b - a) * sd(y) / sqrt(n_samples)
  return(list(estimate = estimate, se = se))
}

# Example: integrate x^2 from 0 to 1 (true answer = 1/3)
result <- monte_carlo_integrate(function(x) x^2, 0, 1, 50000)
cat("Estimate:", result$estimate, "± ", result$se)
cat("\nTrue value: 0.3333...")
cat("\nError:", abs(result$estimate - 1/3))
```

## Importance Sampling

Standard Monte Carlo can be inefficient for functions with most mass in small regions. Importance sampling uses a proposal distribution $q(x)$ that better matches the integrand:

$$\int f(x)dx = \int \frac{f(x)}{q(x)}q(x)dx \approx \frac{1}{N}\sum_{i=1}^N \frac{f(x_i)}{q(x_i)}$$

```{r importance_sampling}
# Importance sampling example
importance_sampling <- function(target_fn, proposal_sampler, proposal_density, 
                              n_samples = 10000) {
  # Generate samples from proposal distribution
  samples <- proposal_sampler(n_samples)
  
  # Compute importance weights
  weights <- target_fn(samples) / proposal_density(samples)
  
  # Estimate integral
  estimate <- mean(weights)
  se <- sd(weights) / sqrt(n_samples)
  
  return(list(estimate = estimate, se = se, samples = samples, weights = weights))
}

# Example: integrate exp(-x^2/2) from 0 to 5 (half of standard normal)
target <- function(x) exp(-x^2/2)
# Use exponential proposal distribution
prop_sampler <- function(n) rexp(n, rate = 1)
prop_density <- function(x) dexp(x, rate = 1)

is_result <- importance_sampling(target, prop_sampler, prop_density, 50000)
cat("Importance sampling estimate:", is_result$estimate)

# Compare with direct Monte Carlo
direct_result <- monte_carlo_integrate(target, 0, 5, 50000)
cat("\nDirect Monte Carlo estimate:", direct_result$estimate)
```

# Markov Chain Monte Carlo (MCMC)

MCMC methods generate samples from the posterior distribution by constructing a Markov chain whose stationary distribution is the target posterior. The key insight is that we don't need to compute the normalizing constant!

## Markov Chain Properties

A Markov chain has the **Markov property**: the next state depends only on the current state, not the entire history.

For MCMC to work, the chain must be:
1. **Irreducible**: Can reach any state from any other state
2. **Aperiodic**: Doesn't get trapped in cycles  
3. **Ergodic**: Combines irreducibility and aperiodicity

## Detailed Balance

The crucial condition for MCMC algorithms:
$$\pi(x)P(x \to y) = \pi(y)P(y \to x)$$

Where $\pi(x)$ is the target distribution and $P(x \to y)$ is the transition probability.

# Metropolis-Hastings Algorithm

The Metropolis-Hastings algorithm is the foundation of most MCMC methods. It satisfies detailed balance by accepting or rejecting proposed moves based on the ratio of posterior probabilities.

## Algorithm Steps

1. **Initialize**: Start with parameter value $\theta^{(0)}$
2. **Propose**: Generate candidate $\theta^* \sim q(\theta^*|\theta^{(t)})$ from proposal distribution
3. **Compute acceptance probability**: 
   $$\alpha = \min\left(1, \frac{P(\theta^*|D)}{P(\theta^{(t)}|D)} \times \frac{q(\theta^{(t)}|\theta^*)}{q(\theta^*|\theta^{(t)})}\right)$$
4. **Accept/Reject**: 
   - If $u \sim \text{Uniform}(0,1) < \alpha$: accept $\theta^{(t+1)} = \theta^*$
   - Otherwise: reject $\theta^{(t+1)} = \theta^{(t)}$
5. **Repeat**: Go to step 2

## Basic Implementation

```{r metropolis_hastings_basic}
# Basic Metropolis-Hastings implementation
metropolis_hastings <- function(log_posterior, initial_value, n_samples, 
                               proposal_sd = 1, burn_in = 1000) {
  
  n_params <- length(initial_value)
  samples <- matrix(NA, n_samples + burn_in, n_params)
  current_value <- initial_value
  current_log_post <- log_posterior(current_value)
  n_accepted <- 0
  
  for (i in 1:(n_samples + burn_in)) {
    # Propose new state
    proposal <- current_value + rnorm(n_params, 0, proposal_sd)
    
    # Compute log posterior for proposal
    proposal_log_post <- log_posterior(proposal)
    
    # Compute acceptance probability (in log space)
    log_alpha <- proposal_log_post - current_log_post
    
    # Accept or reject
    if (log(runif(1)) < log_alpha) {
      current_value <- proposal
      current_log_post <- proposal_log_post
      n_accepted <- n_accepted + 1
    }
    
    samples[i, ] <- current_value
  }
  
  # Remove burn-in
  final_samples <- samples[(burn_in + 1):(n_samples + burn_in), , drop = FALSE]
  
  acceptance_rate <- n_accepted / (n_samples + burn_in)
  
  return(list(
    samples = final_samples,
    acceptance_rate = acceptance_rate,
    all_samples = samples
  ))
}
```

## Example: Bayesian Linear Regression

Let's implement a complete Bayesian analysis of linear regression using MCMC:

```{r bayesian_regression}
# Generate synthetic data
set.seed(123)
n <- 100
x <- seq(0, 10, length.out = n)
true_alpha <- 2.5
true_beta <- 1.2
true_sigma <- 0.8
y <- true_alpha + true_beta * x + rnorm(n, 0, true_sigma)

# Plot data
plot(x, y, pch = 16, col = "steelblue", main = "Synthetic Linear Data")
abline(a = true_alpha, b = true_beta, col = "red", lwd = 2)
legend("topleft", "True relationship", col = "red", lwd = 2)

# Define log posterior (assuming flat priors)
log_posterior_regression <- function(params) {
  alpha <- params[1]
  beta <- params[2] 
  sigma <- params[3]
  
  if (sigma <= 0) return(-Inf)  # Constraint: sigma > 0
  
  # Log likelihood
  predicted <- alpha + beta * x
  log_lik <- sum(dnorm(y, predicted, sigma, log = TRUE))
  
  # Log priors (flat, so contribute 0 in log space)
  log_prior_alpha <- 0  # flat prior
  log_prior_beta <- 0   # flat prior  
  log_prior_sigma <- 0  # flat prior (but constrained > 0)
  
  return(log_lik + log_prior_alpha + log_prior_beta + log_prior_sigma)
}

# Run MCMC
mcmc_result <- metropolis_hastings(
  log_posterior_regression,
  initial_value = c(0, 0, 1),
  n_samples = 10000,
  proposal_sd = c(0.1, 0.05, 0.05),  # Different step sizes per parameter
  burn_in = 2000
)

cat("Acceptance rate:", round(mcmc_result$acceptance_rate, 3))

# Extract samples
alpha_samples <- mcmc_result$samples[, 1]
beta_samples <- mcmc_result$samples[, 2] 
sigma_samples <- mcmc_result$samples[, 3]

# Summary statistics
params_summary <- data.frame(
  Parameter = c("α (intercept)", "β (slope)", "σ (noise)"),
  True_Value = c(true_alpha, true_beta, true_sigma),
  Posterior_Mean = c(mean(alpha_samples), mean(beta_samples), mean(sigma_samples)),
  Posterior_SD = c(sd(alpha_samples), sd(beta_samples), sd(sigma_samples)),
  CI_2.5 = c(quantile(alpha_samples, 0.025), quantile(beta_samples, 0.025), 
             quantile(sigma_samples, 0.025)),
  CI_97.5 = c(quantile(alpha_samples, 0.975), quantile(beta_samples, 0.975), 
              quantile(sigma_samples, 0.975))
)

print(params_summary)
```

## MCMC Diagnostics

### Trace Plots
Trace plots show the evolution of parameter values over iterations:

```{r trace_plots}
par(mfrow = c(3, 1))

# Trace plots
plot(alpha_samples, type = "l", main = "Trace: α (intercept)", 
     ylab = "α", xlab = "Iteration")
abline(h = true_alpha, col = "red", lwd = 2)

plot(beta_samples, type = "l", main = "Trace: β (slope)", 
     ylab = "β", xlab = "Iteration")
abline(h = true_beta, col = "red", lwd = 2)

plot(sigma_samples, type = "l", main = "Trace: σ (noise)", 
     ylab = "σ", xlab = "Iteration")
abline(h = true_sigma, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

### Posterior Distributions

```{r posterior_plots}
par(mfrow = c(2, 2))

# Marginal posteriors
hist(alpha_samples, breaks = 50, probability = TRUE, 
     main = "Posterior: α", xlab = "α", col = "lightblue")
abline(v = true_alpha, col = "red", lwd = 2)
abline(v = mean(alpha_samples), col = "blue", lwd = 2, lty = 2)

hist(beta_samples, breaks = 50, probability = TRUE,
     main = "Posterior: β", xlab = "β", col = "lightgreen") 
abline(v = true_beta, col = "red", lwd = 2)
abline(v = mean(beta_samples), col = "blue", lwd = 2, lty = 2)

hist(sigma_samples, breaks = 50, probability = TRUE,
     main = "Posterior: σ", xlab = "σ", col = "lightcoral")
abline(v = true_sigma, col = "red", lwd = 2)
abline(v = mean(sigma_samples), col = "blue", lwd = 2, lty = 2)

# Joint posterior (α vs β)
plot(alpha_samples, beta_samples, pch = ".", 
     main = "Joint Posterior: α vs β", xlab = "α", ylab = "β")
points(true_alpha, true_beta, col = "red", pch = 16, cex = 2)
points(mean(alpha_samples), mean(beta_samples), col = "blue", pch = 16, cex = 2)

par(mfrow = c(1, 1))
```

### Autocorrelation and Effective Sample Size

```{r autocorrelation}
# Autocorrelation function
compute_autocorr <- function(x, max_lag = 50) {
  n <- length(x)
  autocorr <- numeric(max_lag + 1)
  x_centered <- x - mean(x)
  
  for (k in 0:max_lag) {
    if (k == 0) {
      autocorr[k + 1] <- 1
    } else {
      autocorr[k + 1] <- cor(x_centered[1:(n-k)], x_centered[(k+1):n])
    }
  }
  return(autocorr)
}

# Plot autocorrelation
par(mfrow = c(1, 3))

autocorr_alpha <- compute_autocorr(alpha_samples)
plot(0:50, autocorr_alpha, type = "l", main = "Autocorr: α", 
     xlab = "Lag", ylab = "Autocorrelation")
abline(h = 0, col = "gray")

autocorr_beta <- compute_autocorr(beta_samples) 
plot(0:50, autocorr_beta, type = "l", main = "Autocorr: β",
     xlab = "Lag", ylab = "Autocorrelation")
abline(h = 0, col = "gray")

autocorr_sigma <- compute_autocorr(sigma_samples)
plot(0:50, autocorr_sigma, type = "l", main = "Autocorr: σ",
     xlab = "Lag", ylab = "Autocorrelation") 
abline(h = 0, col = "gray")

par(mfrow = c(1, 1))

# Effective sample size (rough estimate)
eff_sample_size <- function(x) {
  autocorr <- compute_autocorr(x, min(length(x)/4, 200))
  # Find where autocorr first goes negative
  first_negative <- which(autocorr < 0)[1]
  if (is.na(first_negative)) first_negative <- length(autocorr)
  
  tau_int <- 1 + 2 * sum(autocorr[2:first_negative])
  eff_size <- length(x) / (2 * tau_int)
  return(max(1, eff_size))
}

cat("Effective sample sizes:\n")
cat("α:", round(eff_sample_size(alpha_samples)))
cat("\nβ:", round(eff_sample_size(beta_samples))) 
cat("\nσ:", round(eff_sample_size(sigma_samples)))
```

# Advanced MCMC Methods

## Adaptive Metropolis

Adaptive methods tune the proposal distribution during sampling to achieve optimal acceptance rates (typically 20-50%):

```{r adaptive_metropolis}
adaptive_metropolis <- function(log_posterior, initial_value, n_samples,
                               target_rate = 0.3, adaptation_period = 100) {
  
  n_params <- length(initial_value)
  samples <- matrix(NA, n_samples, n_params)
  current_value <- initial_value
  current_log_post <- log_posterior(current_value)
  
  # Initial covariance (scaled identity)
  cov_matrix <- 0.1 * diag(n_params)
  n_accepted <- 0
  
  for (i in 1:n_samples) {
    # Propose new state using current covariance
    proposal <- current_value + mvrnorm(1, rep(0, n_params), cov_matrix)
    proposal_log_post <- log_posterior(proposal)
    
    # Accept/reject
    log_alpha <- proposal_log_post - current_log_post
    if (log(runif(1)) < log_alpha) {
      current_value <- proposal
      current_log_post <- proposal_log_post
      n_accepted <- n_accepted + 1
    }
    
    samples[i, ] <- current_value
    
    # Adapt covariance every adaptation_period iterations
    if (i %% adaptation_period == 0 && i > adaptation_period) {
      recent_acceptance <- n_accepted / i
      
      # Update empirical covariance
      if (i > 2 * n_params) {
        sample_cov <- cov(samples[1:i, ])
        cov_matrix <- 2.38^2 / n_params * sample_cov + 1e-6 * diag(n_params)
      }
      
      # Scale based on acceptance rate
      if (recent_acceptance < target_rate - 0.05) {
        cov_matrix <- cov_matrix * 0.9  # Shrink if too many rejections
      } else if (recent_acceptance > target_rate + 0.05) {
        cov_matrix <- cov_matrix * 1.1  # Expand if too many acceptances
      }
    }
  }
  
  return(list(
    samples = samples,
    acceptance_rate = n_accepted / n_samples,
    final_cov = cov_matrix
  ))
}

# Test adaptive Metropolis on our regression problem
adaptive_result <- adaptive_metropolis(
  log_posterior_regression,
  initial_value = c(0, 0, 1),
  n_samples = 5000,
  target_rate = 0.3
)

cat("Adaptive Metropolis acceptance rate:", round(adaptive_result$acceptance_rate, 3))
```

## Hamiltonian Monte Carlo (HMC)

HMC uses gradient information to propose distant but high-probability states:

```{r hmc_basics}
# Simple HMC implementation (conceptual)
# Note: This is a simplified version for illustration
hamiltonian_mc <- function(log_posterior, grad_log_posterior, initial_value,
                          n_samples, step_size = 0.01, n_steps = 10) {
  
  n_params <- length(initial_value)
  samples <- matrix(NA, n_samples, n_params)
  current_q <- initial_value
  n_accepted <- 0
  
  for (i in 1:n_samples) {
    # Sample momentum
    current_p <- rnorm(n_params, 0, 1)
    
    # Leapfrog integration
    q <- current_q
    p <- current_p
    
    # Half step for momentum
    p <- p + (step_size / 2) * grad_log_posterior(q)
    
    for (j in 1:n_steps) {
      # Full step for position
      q <- q + step_size * p
      
      # Full step for momentum (except last)
      if (j < n_steps) {
        p <- p + step_size * grad_log_posterior(q)
      }
    }
    
    # Half step for momentum
    p <- p + (step_size / 2) * grad_log_posterior(q)
    
    # Negate momentum to make proposal symmetric
    p <- -p
    
    # Compute acceptance probability
    current_H <- -log_posterior(current_q) + 0.5 * sum(current_p^2)
    proposed_H <- -log_posterior(q) + 0.5 * sum(p^2)
    
    log_alpha <- current_H - proposed_H
    
    if (log(runif(1)) < log_alpha) {
      current_q <- q
      n_accepted <- n_accepted + 1
    }
    
    samples[i, ] <- current_q
  }
  
  return(list(
    samples = samples,
    acceptance_rate = n_accepted / n_samples
  ))
}

# For demonstration, we'd need to compute gradients
# This is typically done with automatic differentiation
cat("HMC requires gradient computation - typically implemented in specialized software like Stan")
```

## Gibbs Sampling

When the full conditional distributions are available, Gibbs sampling can be very efficient:

```{r gibbs_sampling}
# Gibbs sampler for bivariate normal
# Demonstrates conditional sampling
gibbs_bivariate_normal <- function(n_samples, mu1 = 0, mu2 = 0, 
                                  sigma1 = 1, sigma2 = 1, rho = 0.5) {
  
  samples <- matrix(NA, n_samples, 2)
  x1 <- 0  # Initialize
  x2 <- 0
  
  for (i in 1:n_samples) {
    # Sample x1 given x2
    mu1_given_x2 <- mu1 + rho * (sigma1 / sigma2) * (x2 - mu2)
    sigma1_given_x2 <- sigma1 * sqrt(1 - rho^2)
    x1 <- rnorm(1, mu1_given_x2, sigma1_given_x2)
    
    # Sample x2 given x1  
    mu2_given_x1 <- mu2 + rho * (sigma2 / sigma1) * (x1 - mu1)
    sigma2_given_x1 <- sigma2 * sqrt(1 - rho^2)
    x2 <- rnorm(1, mu2_given_x1, sigma2_given_x1)
    
    samples[i, ] <- c(x1, x2)
  }
  
  return(samples)
}

# Run Gibbs sampler
gibbs_samples <- gibbs_bivariate_normal(5000, rho = 0.7)

# Compare with direct sampling
direct_samples <- mvrnorm(5000, c(0, 0), matrix(c(1, 0.7, 0.7, 1), 2, 2))

par(mfrow = c(1, 2))
plot(gibbs_samples, main = "Gibbs Sampling", pch = ".", 
     xlab = "X1", ylab = "X2", xlim = c(-4, 4), ylim = c(-4, 4))

plot(direct_samples, main = "Direct Sampling", pch = ".", 
     xlab = "X1", ylab = "X2", xlim = c(-4, 4), ylim = c(-4, 4))
par(mfrow = c(1, 1))
```

# Model Selection and Comparison

Model selection is one of the most challenging aspects of statistical inference. Unlike parameter estimation within a given model, comparing different models requires balancing goodness-of-fit against model complexity. This section covers the major approaches used in practice.

## Theoretical Foundation

### The Bias-Variance Tradeoff

All model selection criteria attempt to balance two competing objectives:

1. **Bias**: Simple models may be too restrictive (underfitting)
2. **Variance**: Complex models may be too flexible (overfitting)

The optimal model minimizes the expected prediction error:
$$\text{EPE} = \text{Bias}^2 + \text{Variance} + \text{Irreducible Error}$$

### Information Theory Perspective

Many criteria are grounded in information theory, specifically Kullback-Leibler (KL) divergence:
$$D_{KL}(f||g) = \int f(x) \log\frac{f(x)}{g(x)} dx$$

This measures the information lost when using model $g$ to approximate true distribution $f$.

## Information Criteria

### Akaike Information Criterion (AIC)

**Developed by**: Hirotugu Akaike (1973)  
**Theoretical basis**: Asymptotic approximation to KL divergence

$$\text{AIC} = 2k - 2\ln(\hat{L})$$

Where:
- $k$ = number of parameters
- $\hat{L}$ = maximum likelihood value

**Key Properties:**
- **Asymptotically unbiased** estimator of KL divergence
- **Consistent** for prediction (selects best predicting model)
- **Not consistent** for model selection (may overfit with large samples)
- Penalty term $2k$ is relatively mild

```{r aic_detailed}
# Detailed AIC implementation and interpretation
compute_aic_detailed <- function(log_likelihood, n_params, n_data) {
  max_log_lik <- max(log_likelihood)
  
  # Standard AIC
  aic <- 2 * n_params - 2 * max_log_lik
  
  # AIC corrected for small samples (AICc)
  if (n_data / n_params < 40) {
    correction <- (2 * n_params * (n_params + 1)) / (n_data - n_params - 1)
    aicc <- aic + correction
  } else {
    aicc <- aic
  }
  
  return(list(
    AIC = aic,
    AICc = aicc,
    penalty = 2 * n_params,
    log_likelihood = max_log_lik,
    small_sample = n_data / n_params < 40
  ))
}

# Example: Compare different polynomial models
set.seed(999)
n_obs <- 50
x_poly <- seq(0, 1, length.out = n_obs)
# True model is quadratic
y_true <- 2 + 3*x_poly - 4*x_poly^2
y_poly <- y_true + rnorm(n_obs, 0, 0.3)

# Fit polynomials of different degrees
max_degree <- 6
aic_comparison <- data.frame(
  degree = 1:max_degree,
  aic = numeric(max_degree),
  aicc = numeric(max_degree),
  bic = numeric(max_degree),
  log_lik = numeric(max_degree),
  n_params = numeric(max_degree)
)

for (degree in 1:max_degree) {
  # Fit polynomial
  X <- poly(x_poly, degree, raw = TRUE)
  model <- lm(y_poly ~ X)
  
  n_params <- length(coef(model))
  log_lik <- logLik(model)[1]
  
  aic_result <- compute_aic_detailed(log_lik, n_params, n_obs)
  
  aic_comparison[degree, ] <- c(
    degree = degree,
    aic = aic_result$AIC,
    aicc = aic_result$AICc, 
    bic = log(n_obs) * n_params - 2 * log_lik,
    log_lik = log_lik,
    n_params = n_params
  )
}

print(aic_comparison)

# Find best models
best_aic <- which.min(aic_comparison$aic)
best_aicc <- which.min(aic_comparison$aicc)
best_bic <- which.min(aic_comparison$bic)

cat("\nBest models:\n")
cat("AIC selects degree:", best_aic, "\n")
cat("AICc selects degree:", best_aicc, "\n") 
cat("BIC selects degree:", best_bic, "\n")
cat("True model degree: 2\n")

# Plot comparison
par(mfrow = c(1, 2))
plot(aic_comparison$degree, aic_comparison$aic, type = "b", 
     main = "Model Selection Criteria", xlab = "Polynomial Degree", 
     ylab = "Information Criterion", pch = 16, col = "red")
lines(aic_comparison$degree, aic_comparison$aicc, type = "b", pch = 17, col = "orange")
lines(aic_comparison$degree, aic_comparison$bic, type = "b", pch = 18, col = "blue")
legend("topright", c("AIC", "AICc", "BIC"), col = c("red", "orange", "blue"), 
       pch = c(16, 17, 18), lty = 1)
abline(v = 2, col = "gray", lty = 2)

# Show the data and selected models
plot(x_poly, y_poly, pch = 16, main = "Data and Selected Models")
lines(x_poly, y_true, col = "black", lwd = 2, lty = 1)

# Plot best AIC model
X_aic <- poly(x_poly, best_aic, raw = TRUE)
model_aic <- lm(y_poly ~ X_aic)
lines(x_poly, predict(model_aic), col = "red", lwd = 2, lty = 2)

# Plot best BIC model  
X_bic <- poly(x_poly, best_bic, raw = TRUE)
model_bic <- lm(y_poly ~ X_bic)
lines(x_poly, predict(model_bic), col = "blue", lwd = 2, lty = 3)

legend("topright", c("True", paste("AIC (deg", best_aic, ")"), paste("BIC (deg", best_bic, ")")), 
       col = c("black", "red", "blue"), lwd = 2, lty = c(1, 2, 3))

par(mfrow = c(1, 1))
```

#### AIC Variants

**AICc (Corrected AIC)**: For small samples where $n/k < 40$:
$$\text{AICc} = \text{AIC} + \frac{2k(k+1)}{n-k-1}$$

**Quasi-AIC (QAIC)**: For overdispersed data:
$$\text{QAIC} = \frac{2k - 2\ln(\hat{L})}{\hat{c}}$$
where $\hat{c}$ is the overdispersion parameter.

### Bayesian Information Criterion (BIC)

**Developed by**: Gideon Schwarz (1978)  
**Also known as**: Schwarz Information Criterion (SIC)  
**Theoretical basis**: Bayesian model selection with flat priors

$$\text{BIC} = k\ln(n) - 2\ln(\hat{L})$$

**Key Properties:**
- **Consistent** for model selection (selects true model as $n \to \infty$)
- **Stronger penalty** than AIC: $k\ln(n)$ vs $2k$
- Approximates **Bayes factor** when priors are uniform
- Favors **simpler models** more than AIC

```{r bic_detailed}
# BIC theoretical foundation and comparison
analyze_bic_penalty <- function() {
  n_values <- c(10, 50, 100, 500, 1000, 5000)
  k_values <- 1:10
  
  penalty_comparison <- expand.grid(n = n_values, k = k_values)
  penalty_comparison$aic_penalty <- 2 * penalty_comparison$k
  penalty_comparison$bic_penalty <- penalty_comparison$k * log(penalty_comparison$n)
  penalty_comparison$ratio <- penalty_comparison$bic_penalty / penalty_comparison$aic_penalty
  
  return(penalty_comparison)
}

penalty_data <- analyze_bic_penalty()

# Plot penalty comparison
library(ggplot2)
ggplot(penalty_data, aes(x = k, y = ratio, color = factor(n))) +
  geom_line(size = 1) +
  geom_point() +
  scale_color_viridis_d(name = "Sample Size (n)") +
  labs(title = "BIC vs AIC Penalty Ratio",
       subtitle = "BIC becomes increasingly harsh relative to AIC as sample size grows",
       x = "Number of Parameters (k)", 
       y = "BIC Penalty / AIC Penalty") +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  theme_minimal()

# Show crossover points
crossover_points <- penalty_data[penalty_data$ratio > 1, ]
min_crossover <- min(crossover_points$n)
cat("BIC penalty exceeds AIC when n ≥", min_crossover, "for k=1\n")
cat("At n=100, BIC penalty is", round(log(100)/2, 2), "times AIC penalty\n")
```

**Relationship to Bayes Factors:**

BIC approximates $-2\log(BF_{01})$ where $BF_{01}$ is the Bayes factor comparing models 0 and 1:
$$BF_{01} \approx \exp\left(\frac{\text{BIC}_1 - \text{BIC}_0}{2}\right)$$

This makes BIC differences interpretable:
- $\Delta\text{BIC} < 2$: Weak evidence
- $2 < \Delta\text{BIC} < 6$: Positive evidence  
- $6 < \Delta\text{BIC} < 10$: Strong evidence
- $\Delta\text{BIC} > 10$: Very strong evidence

### Deviance Information Criterion (DIC)

**Developed by**: Spiegelhalter et al. (2002)  
**Context**: Bayesian analysis with MCMC samples  
**Advantage**: Uses full posterior distribution, not just point estimates

$$\text{DIC} = \bar{D} + p_D = D(\bar{\theta}) + 2p_D$$

Where:
- $D(\theta) = -2\log p(y|\theta)$ (deviance)
- $\bar{D} = E_{\theta|y}[D(\theta)]$ (posterior mean deviance)
- $D(\bar{\theta}) = D(E_{\theta|y}[\theta])$ (deviance at posterior mean)
- $p_D = \bar{D} - D(\bar{\theta})$ (effective number of parameters)

```{r dic_detailed}
# Comprehensive DIC implementation
compute_dic_comprehensive <- function(mcmc_samples, log_likelihood_fn, data) {
  n_samples <- nrow(mcmc_samples)
  n_params <- ncol(mcmc_samples)
  
  # Compute deviance for each MCMC sample
  deviances <- numeric(n_samples)
  log_likelihoods <- numeric(n_samples)
  
  for (i in 1:n_samples) {
    log_lik <- log_likelihood_fn(mcmc_samples[i, ], data)
    log_likelihoods[i] <- log_lik
    deviances[i] <- -2 * log_lik
  }
  
  # Posterior mean deviance
  D_bar <- mean(deviances)
  
  # Deviance at posterior mean parameters
  theta_bar <- colMeans(mcmc_samples)
  D_theta_bar <- -2 * log_likelihood_fn(theta_bar, data)
  
  # Effective number of parameters
  p_D <- D_bar - D_theta_bar
  
  # DIC
  dic <- D_bar + p_D
  
  # Alternative formulation: DIC = D(theta_bar) + 2*p_D
  dic_alt <- D_theta_bar + 2 * p_D
  
  # Variance-based effective parameters
  p_V <- var(deviances) / 2
  dic_v <- D_bar + p_V
  
  return(list(
    DIC = dic,
    DIC_alt = dic_alt,
    DIC_variance = dic_v,
    D_bar = D_bar,
    D_theta_bar = D_theta_bar,
    p_D = p_D,
    p_V = p_V,
    effective_params_ratio = p_D / n_params,
    deviances = deviances,
    log_likelihoods = log_likelihoods
  ))
}

# Apply to our regression example
regression_log_lik <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  sigma <- params[3]
  
  predicted <- alpha + beta * data$x
  return(sum(dnorm(data$y, predicted, sigma, log = TRUE)))
}

data_list <- list(x = x, y = y)
dic_result <- compute_dic_comprehensive(mcmc_result$samples, regression_log_lik, data_list)

cat("DIC Analysis Results:\n")
cat("DIC =", round(dic_result$DIC, 2), "\n")
cat("Effective parameters (p_D) =", round(dic_result$p_D, 2), "\n")
cat("Actual parameters =", ncol(mcmc_result$samples), "\n")
cat("Effective/Actual ratio =", round(dic_result$effective_params_ratio, 3), "\n")

# Plot deviance distribution
par(mfrow = c(1, 2))
hist(dic_result$deviances, breaks = 30, probability = TRUE,
     main = "Posterior Distribution of Deviance", 
     xlab = "Deviance", col = "lightblue")
abline(v = dic_result$D_bar, col = "red", lwd = 2)
abline(v = dic_result$D_theta_bar, col = "blue", lwd = 2)
legend("topright", c("Mean Deviance", "Deviance at Mean"), 
       col = c("red", "blue"), lwd = 2)

# Plot effective parameters over iterations
effective_params_running <- cumsum(dic_result$deviances) / (1:length(dic_result$deviances)) - 
                           dic_result$D_theta_bar
plot(effective_params_running, type = "l", 
     main = "Running Estimate of Effective Parameters",
     xlab = "MCMC Iteration", ylab = "p_D")
abline(h = dic_result$p_D, col = "red", lwd = 2)
abline(h = ncol(mcmc_result$samples), col = "blue", lwd = 2, lty = 2)
legend("topright", c("Final p_D", "Actual parameters"), 
       col = c("red", "blue"), lwd = 2, lty = c(1, 2))

par(mfrow = c(1, 1))
```

**Interpretation of Effective Parameters:**
- $p_D \approx k$: Parameters are well-identified
- $p_D < k$: Parameters are constrained by priors or correlations
- $p_D > k$: Model may be overfitting or have identification issues

**Limitations of DIC:**
- Can give negative $p_D$ (problematic)
- Not appropriate for model averaging
- Can be unstable for mixture models
- Assumes posterior is approximately normal

### Watanabe-Akaike Information Criterion (WAIC)

**Developed by**: Sumio Watanabe (2010)  
**Advantages**: More robust than DIC, pointwise calculation  
**Also known as**: Widely Applicable Information Criterion

$$\text{WAIC} = -2(\text{lppd} - p_{\text{WAIC}})$$

Where:
- $\text{lppd} = \sum_{i=1}^n \log\left(\frac{1}{S}\sum_{s=1}^S p(y_i|\theta^s)\right)$ (log pointwise predictive density)
- $p_{\text{WAIC}} = \sum_{i=1}^n \text{Var}_s(\log p(y_i|\theta^s))$ (effective parameters)

```{r waic_detailed}
# WAIC implementation
compute_waic <- function(mcmc_samples, log_likelihood_pointwise_fn, data) {
  n_samples <- nrow(mcmc_samples)
  n_data <- length(data$y)
  
  # Matrix to store pointwise log-likelihoods
  log_lik_matrix <- matrix(NA, n_samples, n_data)
  
  # Compute pointwise log-likelihoods for each MCMC sample
  for (s in 1:n_samples) {
    log_lik_matrix[s, ] <- log_likelihood_pointwise_fn(mcmc_samples[s, ], data)
  }
  
  # Compute lppd (log pointwise predictive density)
  lppd_terms <- numeric(n_data)
  for (i in 1:n_data) {
    # Log of average likelihood for observation i
    lppd_terms[i] <- log(mean(exp(log_lik_matrix[, i])))
  }
  lppd <- sum(lppd_terms)
  
  # Compute effective parameters (penalty term)
  p_waic_terms <- numeric(n_data)
  for (i in 1:n_data) {
    p_waic_terms[i] <- var(log_lik_matrix[, i])
  }
  p_waic <- sum(p_waic_terms)
  
  # WAIC
  waic <- -2 * (lppd - p_waic)
  
  # Standard error of WAIC
  waic_se <- sqrt(n_data * var(-2 * (lppd_terms - p_waic_terms)))
  
  return(list(
    WAIC = waic,
    lppd = lppd,
    p_waic = p_waic,
    pointwise_waic = -2 * (lppd_terms - p_waic_terms),
    waic_se = waic_se,
    log_lik_matrix = log_lik_matrix
  ))
}

# Pointwise log-likelihood function
pointwise_log_lik <- function(params, data) {
  alpha <- params[1]
  beta <- params[2] 
  sigma <- params[3]
  
  predicted <- alpha + beta * data$x
  return(dnorm(data$y, predicted, sigma, log = TRUE))
}

waic_result <- compute_waic(mcmc_result$samples, pointwise_log_lik, data_list)

cat("WAIC Analysis Results:\n")
cat("WAIC =", round(waic_result$WAIC, 2), "±", round(waic_result$waic_se, 2), "\n")
cat("lppd =", round(waic_result$lppd, 2), "\n") 
cat("Effective parameters (p_WAIC) =", round(waic_result$p_waic, 2), "\n")

# Compare all criteria
all_criteria <- data.frame(
  Criterion = c("AIC", "BIC", "DIC", "WAIC"),
  Value = c(
    aic_comparison[2, "aic"],  # Use degree 2 from polynomial example
    aic_comparison[2, "bic"],
    dic_result$DIC,
    waic_result$WAIC
  ),
  Effective_Params = c(
    2,  # Fixed for AIC/BIC
    2,
    dic_result$p_D,
    waic_result$p_waic
  )
)

print(all_criteria)

# Plot pointwise WAIC contributions
plot(waic_result$pointwise_waic, type = "p", pch = 16,
     main = "Pointwise WAIC Contributions", 
     xlab = "Observation", ylab = "Pointwise WAIC")
abline(h = mean(waic_result$pointwise_waic), col = "red", lwd = 2)
```

### Leave-One-Out Cross-Validation (LOO-CV)

**Theoretical optimum**: Uses all available data for both training and validation  
**Challenge**: Computationally expensive ($n$ model fits)  
**PSIS-LOO**: Pareto Smoothed Importance Sampling approximation

```{r loo_cv_detailed}
# Approximate LOO using importance sampling
compute_loo_is <- function(log_lik_matrix) {
  n_samples <- nrow(log_lik_matrix)
  n_data <- ncol(log_lik_matrix)
  
  loo_pointwise <- numeric(n_data)
  pareto_k <- numeric(n_data)
  
  for (i in 1:n_data) {
    # Raw importance weights (leave-one-out)
    log_weights <- -log_lik_matrix[, i]
    
    # Normalize weights
    max_log_weight <- max(log_weights)
    log_weights_norm <- log_weights - max_log_weight
    weights <- exp(log_weights_norm)
    weights <- weights / sum(weights)
    
    # Estimate LOO likelihood  
    loo_pointwise[i] <- log(sum(weights * exp(log_lik_matrix[, i])))
    
    # Diagnose with Pareto k (simplified)
    # In practice, would use proper Pareto smoothing
    if (max(weights) > 0.7) {
      pareto_k[i] <- 1  # High k indicates unreliable estimate
    } else {
      pareto_k[i] <- 0.5  # Reasonable estimate
    }
  }
  
  loo_total <- sum(loo_pointwise)
  
  # Effective parameters
  p_loo <- waic_result$lppd - loo_total
  
  # LOO-IC  
  looic <- -2 * loo_total
  
  return(list(
    LOO = loo_total,
    LOOIC = looic,
    p_loo = p_loo,
    pointwise_loo = loo_pointwise,
    pareto_k = pareto_k,
    reliable = mean(pareto_k < 0.7)
  ))
}

loo_result <- compute_loo_is(waic_result$log_lik_matrix)

cat("LOO-CV Results:\n")
cat("LOO =", round(loo_result$LOO, 2), "\n")
cat("LOOIC =", round(loo_result$LOOIC, 2), "\n")
cat("p_loo =", round(loo_result$p_loo, 2), "\n")
cat("Reliable estimates:", round(100 * loo_result$reliable, 1), "%\n")

# Compare WAIC and LOO
comparison_df <- data.frame(
  Method = c("WAIC", "LOO-CV"),
  IC_Value = c(waic_result$WAIC, loo_result$LOOIC),
  Effective_Params = c(waic_result$p_waic, loo_result$p_loo)
)
print(comparison_df)
```

## Advanced Model Selection Topics

### Model Averaging

Instead of selecting a single "best" model, model averaging weights predictions by model probabilities:

$$\hat{y} = \sum_{m=1}^M w_m \hat{y}_m$$

where $w_m$ are model weights (e.g., from AIC weights).

```{r model_averaging}
# AIC model averaging example
compute_aic_weights <- function(aic_values) {
  delta_aic <- aic_values - min(aic_values)
  weights <- exp(-0.5 * delta_aic)
  return(weights / sum(weights))
}

# Use polynomial comparison from earlier
aic_weights <- compute_aic_weights(aic_comparison$aic)

cat("AIC Model Weights:\n")
for (i in 1:max_degree) {
  cat("Degree", i, ":", round(aic_weights[i], 3), "\n")
}

# Model-averaged prediction
x_new <- seq(0, 1, length.out = 100)
pred_averaged <- numeric(length(x_new))

for (degree in 1:max_degree) {
  X_pred <- poly(x_new, degree, raw = TRUE)
  X_fit <- poly(x_poly, degree, raw = TRUE)
  model <- lm(y_poly ~ X_fit)
  
  # Predict (need to be careful with poly() predictions)
  pred_degree <- predict(model, newdata = data.frame(X_fit = X_pred))
  pred_averaged <- pred_averaged + aic_weights[degree] * pred_degree
}

# Plot model averaging result
plot(x_poly, y_poly, pch = 16, main = "Model Averaging vs Best Model")
lines(x_new, 2 + 3*x_new - 4*x_new^2, col = "black", lwd = 2)  # True
lines(x_new, pred_averaged, col = "purple", lwd = 2, lty = 2)   # Averaged
legend("topright", c("True", "Model Averaged"), 
       col = c("black", "purple"), lwd = 2, lty = c(1, 2))
```

### Cross-Validation Strategies

```{r cv_strategies}
# K-fold cross-validation
k_fold_cv <- function(data_x, data_y, k = 5, max_degree = 6) {
  n <- length(data_y)
  fold_size <- floor(n / k)
  cv_errors <- matrix(NA, k, max_degree)
  
  for (fold in 1:k) {
    # Create test indices
    start_idx <- (fold - 1) * fold_size + 1
    end_idx <- min(fold * fold_size, n)
    test_idx <- start_idx:end_idx
    train_idx <- setdiff(1:n, test_idx)
    
    # Split data
    x_train <- data_x[train_idx]
    y_train <- data_y[train_idx]
    x_test <- data_x[test_idx]
    y_test <- data_y[test_idx]
    
    for (degree in 1:max_degree) {
      # Fit on training data
      X_train <- poly(x_train, degree, raw = TRUE)
      model <- lm(y_train ~ X_train)
      
      # Predict on test data
      X_test <- poly(x_test, degree, raw = TRUE)
      pred_test <- predict(model, newdata = data.frame(X_train = X_test))
      
      # Compute prediction error
      cv_errors[fold, degree] <- mean((y_test - pred_test)^2)
    }
  }
  
  return(colMeans(cv_errors))
}

cv_errors <- k_fold_cv(x_poly, y_poly, k = 5)
best_cv <- which.min(cv_errors)

cat("Cross-validation selects degree:", best_cv, "\n")

# Plot CV errors
plot(1:max_degree, cv_errors, type = "b", pch = 16,
     main = "5-Fold Cross-Validation Error",
     xlab = "Polynomial Degree", ylab = "CV Error")
abline(v = best_cv, col = "red", lty = 2)
abline(v = 2, col = "gray", lty = 2)  # True model
```

## Practical Guidelines

### Which Criterion to Use?

| **Criterion** | **Best for** | **Avoid when** |
|---------------|--------------|----------------|
| **AIC** | Prediction, model averaging | Want parsimonious model |
| **BIC** | Model selection, large samples | Small samples, prediction focus |
| **DIC** | Bayesian models, MCMC | Mixture models, non-normal posteriors |
| **WAIC** | General Bayesian analysis | Computational constraints |
| **LOO-CV** | Gold standard | Very large datasets |

### Sample Size Considerations

```{r sample_size_effects}
# Demonstrate how criteria behave with different sample sizes
simulate_model_selection <- function(n, true_degree = 2, noise_sd = 0.3) {
  x_sim <- seq(0, 1, length.out = n)
  y_sim <- 2 + 3*x_sim - 4*x_sim^2 + rnorm(n, 0, noise_sd)
  
  results <- data.frame(degree = 1:5, aic = NA, bic = NA)
  
  for (deg in 1:5) {
    X_sim <- poly(x_sim, deg, raw = TRUE)
    model <- lm(y_sim ~ X_sim)
    
    log_lik <- logLik(model)[1]
    n_params <- length(coef(model))
    
    results$aic[deg] <- 2 * n_params - 2 * log_lik
    results$bic[deg] <- log(n) * n_params - 2 * log_lik
  }
  
  return(list(
    aic_best = which.min(results$aic),
    bic_best = which.min(results$bic),
    results = results
  ))
}

# Test different sample sizes
sample_sizes <- c(20, 50, 100, 200, 500)
n_simulations <- 100

selection_results <- data.frame(
  n = rep(sample_sizes, each = n_simulations),
  aic_correct = logical(length(sample_sizes) * n_simulations),
  bic_correct = logical(length(sample_sizes) * n_simulations)
)

set.seed(123)
for (i in seq_along(sample_sizes)) {
  for (sim in 1:n_simulations) {
    idx <- (i - 1) * n_simulations + sim
    result <- simulate_model_selection(sample_sizes[i])
    
    selection_results$aic_correct[idx] <- result$aic_best == 2
    selection_results$bic_correct[idx] <- result$bic_best == 2
  }
}

# Summarize results
accuracy_summary <- aggregate(cbind(aic_correct, bic_correct) ~ n, 
                             data = selection_results, FUN = mean)

print(accuracy_summary)

# Plot accuracy vs sample size
plot(accuracy_summary$n, accuracy_summary$aic_correct, type = "b", 
     col = "red", pch = 16, ylim = c(0, 1),
     main = "Model Selection Accuracy vs Sample Size",
     xlab = "Sample Size", ylab = "Proportion Correct")
lines(accuracy_summary$n, accuracy_summary$bic_correct, type = "b", 
      col = "blue", pch = 17)
legend("bottomright", c("AIC", "BIC"), col = c("red", "blue"), 
       pch = c(16, 17), lty = 1)
```

## Leave-One-Out Cross-Validation (LOO)

```{r loo_cv}
# Approximate LOO using Pareto smoothed importance sampling
# This is a simplified implementation
compute_loo <- function(log_lik_matrix) {
  n_data <- ncol(log_lik_matrix)
  n_samples <- nrow(log_lik_matrix)
  
  loo_estimates <- numeric(n_data)
  
  for (i in 1:n_data) {
    # Importance weights for leaving out observation i
    # This is a simplified version - real implementation is more complex
    weights <- exp(log_lik_matrix[, i])
    loo_estimates[i] <- log(mean(weights))
  }
  
  total_loo <- sum(loo_estimates)
  return(list(total_loo = total_loo, pointwise = loo_estimates))
}

cat("LOO computation requires careful implementation - typically use 'loo' package in R")
```

# Practical Guidelines and Best Practices

## Tuning MCMC

### Acceptance Rates
- **Metropolis-Hastings**: Aim for 20-50% acceptance rate
- **HMC**: Aim for 60-90% acceptance rate  
- **Random Walk Metropolis**: ~23% for high dimensions

### Proposal Distributions
```{r proposal_tuning}
# Test different proposal standard deviations
test_proposals <- function(log_posterior, true_values, proposal_sds) {
  results <- data.frame(
    proposal_sd = proposal_sds,
    acceptance_rate = numeric(length(proposal_sds)),
    efficiency = numeric(length(proposal_sds))
  )
  
  for (i in seq_along(proposal_sds)) {
    result <- metropolis_hastings(
      log_posterior, 
      initial_value = true_values,
      n_samples = 2000,
      proposal_sd = proposal_sds[i],
      burn_in = 500
    )
    
    results$acceptance_rate[i] <- result$acceptance_rate
    
    # Rough efficiency measure (inverse of autocorr time)
    ess <- eff_sample_size(result$samples[, 1])
    results$efficiency[i] <- ess / nrow(result$samples)
  }
  
  return(results)
}

# Test range of proposal SDs
proposal_test <- test_proposals(
  log_posterior_regression,
  c(true_alpha, true_beta, true_sigma),
  seq(0.01, 0.5, by = 0.05)
)

par(mfrow = c(1, 2))
plot(proposal_test$proposal_sd, proposal_test$acceptance_rate, 
     type = "b", main = "Acceptance Rate vs Proposal SD",
     xlab = "Proposal SD", ylab = "Acceptance Rate")
abline(h = c(0.2, 0.5), col = "red", lty = 2)

plot(proposal_test$proposal_sd, proposal_test$efficiency,
     type = "b", main = "Efficiency vs Proposal SD", 
     xlab = "Proposal SD", ylab = "Effective Sample Size Ratio")
par(mfrow = c(1, 1))
```

## Convergence Diagnostics

### Gelman-Rubin Statistic (R-hat)

```{r convergence_diagnostics}
# Multiple chains for convergence assessment
run_multiple_chains <- function(n_chains = 4) {
  chains <- vector("list", n_chains)
  
  for (i in 1:n_chains) {
    # Start from different initial values
    initial <- c(runif(1, 0, 5), runif(1, 0, 2), runif(1, 0.5, 1.5))
    
    chains[[i]] <- metropolis_hastings(
      log_posterior_regression,
      initial_value = initial,
      n_samples = 2500,
      proposal_sd = c(0.1, 0.05, 0.05),
      burn_in = 500
    )$samples
  }
  
  return(chains)
}

# Run multiple chains
multiple_chains <- run_multiple_chains(4)

# Simple R-hat computation
compute_rhat <- function(chains) {
  n_chains <- length(chains)
  n_samples <- nrow(chains[[1]])
  n_params <- ncol(chains[[1]])
  
  rhat_values <- numeric(n_params)
  
  for (p in 1:n_params) {
    # Extract parameter p from all chains
    chain_data <- sapply(chains, function(x) x[, p])
    
    # Between-chain variance
    chain_means <- colMeans(chain_data)
    overall_mean <- mean(chain_means)
    B <- n_samples * var(chain_means)
    
    # Within-chain variance
    W <- mean(apply(chain_data, 2, var))
    
    # Estimated variance
    var_est <- ((n_samples - 1) / n_samples) * W + (1 / n_samples) * B
    
    # R-hat
    rhat_values[p] <- sqrt(var_est / W)
  }
  
  return(rhat_values)
}

rhat_vals <- compute_rhat(multiple_chains)
param_names <- c("α", "β", "σ")

for (i in 1:3) {
  cat(sprintf("R-hat for %s: %.3f", param_names[i], rhat_vals[i]))
  if (rhat_vals[i] < 1.1) cat(" ✓") else cat(" ⚠")
  cat("\n")
}
```

### Trace Plot Examination

```{r trace_diagnostics}
# Plot traces from multiple chains
par(mfrow = c(3, 1))
colors <- c("red", "blue", "green", "purple")

for (p in 1:3) {
  plot(1, type = "n", xlim = c(1, 2500), 
       ylim = range(sapply(multiple_chains, function(x) range(x[, p]))),
       main = paste("Trace Plot:", param_names[p]),
       xlab = "Iteration", ylab = param_names[p])
  
  for (chain in 1:4) {
    lines(multiple_chains[[chain]][, p], col = colors[chain], alpha = 0.7)
  }
  
  abline(h = c(true_alpha, true_beta, true_sigma)[p], col = "black", lwd = 2, lty = 2)
}

par(mfrow = c(1, 1))
```

# Advanced Topics

## Hierarchical Models

```{r hierarchical_example}
# Example: Hierarchical normal model
# y_ij ~ N(theta_i, sigma^2)  [data level]
# theta_i ~ N(mu, tau^2)      [group level] 
# mu ~ N(0, 100)              [hyperprior]
# tau ~ Half-Cauchy(0, 5)     [hyperprior]
# sigma ~ Half-Cauchy(0, 5)   [hyperprior]

# Generate hierarchical data
set.seed(456)
n_groups <- 5
n_per_group <- 10
true_mu <- 5
true_tau <- 2
true_sigma <- 1

true_theta <- rnorm(n_groups, true_mu, true_tau)
hierarchical_data <- vector("list", n_groups)

for (i in 1:n_groups) {
  hierarchical_data[[i]] <- rnorm(n_per_group, true_theta[i], true_sigma)
}

# Flatten data for analysis
y_flat <- unlist(hierarchical_data)
group_id <- rep(1:n_groups, each = n_per_group)

# Log posterior for hierarchical model
log_posterior_hierarchical <- function(params) {
  mu <- params[1]
  log_tau <- params[2]
  log_sigma <- params[3] 
  theta <- params[4:(3 + n_groups)]
  
  tau <- exp(log_tau)
  sigma <- exp(log_sigma)
  
  if (tau <= 0 || sigma <= 0) return(-Inf)
  
  # Data likelihood
  log_lik <- 0
  for (i in 1:n_groups) {
    group_data <- hierarchical_data[[i]]
    log_lik <- log_lik + sum(dnorm(group_data, theta[i], sigma, log = TRUE))
  }
  
  # Group-level prior
  log_prior_theta <- sum(dnorm(theta, mu, tau, log = TRUE))
  
  # Hyperpriors
  log_prior_mu <- dnorm(mu, 0, 10, log = TRUE)
  log_prior_tau <- dhalfcauchy_log(tau, 5)  # Would need custom function
  log_prior_sigma <- dhalfcauchy_log(sigma, 5)  # Would need custom function
  
  return(log_lik + log_prior_theta + log_prior_mu + log_prior_tau + log_prior_sigma)
}

cat("Hierarchical models require careful implementation of hyperpriors")
cat("\nTypically implemented in specialized software like Stan, BUGS, or JAGS")
```

## Mixture Models

```{r mixture_models}
# Simple 2-component Gaussian mixture
# Demonstrates label switching problem

# Generate mixture data
set.seed(789)
n1 <- 100
n2 <- 150
true_mu1 <- -2
true_mu2 <- 3
true_sigma1 <- 1
true_sigma2 <- 1.5
true_pi <- 0.4  # mixing proportion

component1_data <- rnorm(n1, true_mu1, true_sigma1)
component2_data <- rnorm(n2, true_mu2, true_sigma2)
mixture_data <- c(component1_data, component2_data)

# Plot mixture data
hist(mixture_data, breaks = 30, probability = TRUE, 
     main = "Mixture Data", xlab = "Value", col = "lightblue")

# Overlay true components
x_seq <- seq(min(mixture_data), max(mixture_data), length.out = 200)
true_density <- true_pi * dnorm(x_seq, true_mu1, true_sigma1) + 
                (1 - true_pi) * dnorm(x_seq, true_mu2, true_sigma2)
lines(x_seq, true_density, col = "red", lwd = 2)

cat("Mixture models suffer from label switching - identifiability issues")
cat("\nRequire specialized techniques or constraints for proper inference")
```

# Software and Tools

## R Packages

### Essential MCMC Packages
- **`coda`**: MCMC diagnostics and analysis
- **`MCMCpack`**: Ready-made MCMC functions
- **`rstan`**: Interface to Stan (HMC/NUTS)
- **`rjags`**: Interface to JAGS (Gibbs sampling)
- **`LaplacesDemon`**: Comprehensive Bayesian computing

### Example with `coda` package

```{r coda_example}
library(coda)

# Convert our MCMC samples to coda format
mcmc_coda <- mcmc(mcmc_result$samples)

# Diagnostic plots
# plot(mcmc_coda)  # Would show trace and density plots

# Summary statistics
summary(mcmc_coda)

# Effective sample size
effectiveSize(mcmc_coda)

# Autocorrelation
autocorr.plot(mcmc_coda, lag.max = 50)
```

## Stan Example

```{r stan_example}
# Stan model specification (would be in separate .stan file)
stan_model_code <- "
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  // Priors
  alpha ~ normal(0, 10);
  beta ~ normal(0, 10);
  sigma ~ cauchy(0, 5);
  
  // Likelihood
  y ~ normal(alpha + beta * x, sigma);
}
"

# This would run Stan if available:
# library(rstan)
# stan_fit <- stan(model_code = stan_model_code, 
#                  data = list(N = length(x), x = x, y = y),
#                  chains = 4, iter = 2000)

cat("Stan provides state-of-the-art HMC/NUTS sampling")
cat("\nAutomatic differentiation and adaptive tuning")
```

# Summary and Recommendations

## When to Use Different Methods

### Metropolis-Hastings
- **Good for**: Simple problems, learning MCMC concepts
- **Pros**: Easy to implement, general purpose
- **Cons**: Can be inefficient, requires tuning

### Gibbs Sampling  
- **Good for**: Problems with known conditional distributions
- **Pros**: No tuning required, often efficient
- **Cons**: Limited applicability, can be slow for correlated parameters

### Hamiltonian Monte Carlo (HMC/NUTS)
- **Good for**: Complex, high-dimensional problems
- **Pros**: Very efficient, automatic tuning (NUTS)
- **Cons**: Requires gradients, complex implementation

### Adaptive Methods
- **Good for**: When optimal proposal distribution unknown
- **Pros**: Self-tuning, often more efficient
- **Cons**: More complex, convergence theory more involved

## Best Practices Checklist

1. **Start Simple**: Begin with basic Metropolis-Hastings
2. **Check Convergence**: Use multiple chains, R-hat < 1.1
3. **Examine Traces**: Look for good mixing, no trends
4. **Assess Autocorrelation**: Thin if necessary
5. **Validate**: Compare with known solutions when possible
6. **Use Modern Software**: Stan, JAGS for complex problems

## Common Pitfalls

1. **Insufficient Burn-in**: Not allowing chain to reach stationary distribution
2. **Poor Mixing**: Proposal distribution too narrow or wide
3. **Ignoring Correlations**: Not accounting for parameter correlations
4. **Label Switching**: In mixture models, modes may switch
5. **Improper Priors**: Can lead to improper posteriors
6. **IC Misinterpretation**: Treating IC differences as hypothesis tests
7. **Model Selection Bias**: Using same data for selection and inference

## Information Criteria Pitfalls

### Common Mistakes

**AIC Misuse:**
- Comparing non-nested models inappropriately
- Ignoring AICc correction for small samples
- Using AIC for causal inference

**BIC Limitations:**
- Assumes one true model exists
- Very sensitive to sample size
- May be too conservative for prediction

**DIC Problems:**
- Negative effective parameters
- Instability with mixture models
- Assumes unimodal posteriors

**General IC Issues:**
- Model set dependence (adding bad models can change rankings)
- Ignores model uncertainty
- Not appropriate for causal inference

```{r ic_pitfalls_demo}
# Demonstrate model set dependence
set.seed(555)
n_demo <- 100
x_demo <- runif(n_demo, 0, 10)
y_demo <- 2 + 0.5 * x_demo + rnorm(n_demo, 0, 1)

# Model set 1: Simple models
models_set1 <- list(
  "Intercept only" = lm(y_demo ~ 1),
  "Linear" = lm(y_demo ~ x_demo),
  "Quadratic" = lm(y_demo ~ x_demo + I(x_demo^2))
)

# Model set 2: Include bad complex model
models_set2 <- list(
  "Intercept only" = lm(y_demo ~ 1),
  "Linear" = lm(y_demo ~ x_demo),
  "Quadratic" = lm(y_demo ~ x_demo + I(x_demo^2)),
  "High polynomial" = lm(y_demo ~ poly(x_demo, 8))
)

compute_model_ics <- function(model_list) {
  results <- data.frame(
    Model = names(model_list),
    AIC = sapply(model_list, AIC),
    BIC = sapply(model_list, BIC)
  )
  results$Delta_AIC <- results$AIC - min(results$AIC)
  results$Delta_BIC <- results$BIC - min(results$BIC)
  results$AIC_Weight <- exp(-0.5 * results$Delta_AIC)
  results$AIC_Weight <- results$AIC_Weight / sum(results$AIC_Weight)
  return(results)
}

cat("Model Set 1 (Simple models):\n")
print(compute_model_ics(models_set1))

cat("\nModel Set 2 (With complex model):\n") 
print(compute_model_ics(models_set2))

cat("\nNote how AIC weights change when bad model is added!\n")
```

## Computational Considerations

- **Parallel Chains**: Run multiple chains in parallel
- **Vectorization**: Use vectorized operations when possible  
- **Memory Management**: Store only necessary samples
- **Monitoring**: Track acceptance rates and ESS during sampling

This comprehensive guide covers the essential aspects of Bayesian computation and MCMC methods. The key insight is that MCMC allows us to sample from complex posterior distributions without computing intractable normalizing constants, enabling sophisticated statistical inference in realistic problems.

---

**Further Reading:**
- Gelman et al. "Bayesian Data Analysis" 
- Robert & Casella "Monte Carlo Statistical Methods"
- Neal "MCMC Using Hamiltonian Dynamics"
- Stan User's Guide and Reference Manual
